---
title: "Data Cleaning"
author: "Zachary Clement"
date: "12/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm using files downloaded from the (NCES)[https://nces.ed.gov/programs/digest/2019menu_tables.asp]

First, let's see if we can extract the field of study from each excel sheet
```{r}
for (file in list.files('gender_excel_files')){
  current_sex_df = readxl::read_xls(paste0('gender_excel_files/', file))
  
  id_string = colnames(current_sex_df)[1]
  
  start_removed = gsub(pattern = ".{1,} Degrees in ", x = id_string, replacement = '' )
  end_removed = gsub(pattern = " conferred .{1,}" ,x= start_removed, replacement = '')
  
  print(end_removed)

}
```

Now, let's see if the columns are consistent between sheets
```{r}
num_cols = c()
for (file in list.files('gender_excel_files')){
  current_sex_df = readxl::read_xls(paste0('gender_excel_files/', file))

  num_cols = c(num_cols, dim(current_sex_df)[2])

}
print(num_cols)
```

Looks like the number of columns are not consistent between excel sheets. Let's look at which files are different 

```{r}
num_cols = c()
for (file in list.files('gender_excel_files')){
  current_sex_df = readxl::read_xls(paste0('gender_excel_files/', file))
  print(file)
  print(dim(current_sex_df)[2])


}

```

Looks like I downloaded some files that I shouldn't have. I had assumed that all tables in that list were by sex, but some weren't. I'll delete those and try again. 

```{r}
num_cols = c()
for (file in list.files('gender_excel_files')){
  current_sex_df = readxl::read_xls(paste0('gender_excel_files/', file), skip = 4)
  print(colnames(current_sex_df))
  print(file)
  print(dim(current_sex_df)[2])
}

```

It looks like for some reason, we're imputing another column where not needed on those sheets. Let's just take the first 12 columns of these sheets and see how that looks. 

```{r}
colnames_sex_df = c("year", "bach_t", "bach_annual_pct_change",
                    "bach_m", "bach_f", "bach_f_pct", "mast_t","mast_m",
                    'mast_f', 'doct_t', 'doct_m', 'doct_f')

rm("fields_by_sex") #remove before starting so we can re-run the chunk
for (file in list.files('gender_excel_files')){
  current_sex_df = readxl::read_xls(paste0('gender_excel_files/', file))
  id_string = colnames(current_sex_df)[1]
  current_sex_df = current_sex_df[-(1:4),1:12] #remove 13th col if exists, remove first 4 rows because they're just column descriptions
  # get field name
  colnames(current_sex_df) = colnames_sex_df 
    start_removed = gsub(pattern = ".{1,} Degrees in ", x = id_string, replacement = '' )
  end_removed = gsub(pattern = " conferred .{1,}" ,x= start_removed, replacement = '')
  
  
  ## remove any useless rows from each df
  keep_rows = grep(pattern = "^\\d{4}-\\d{2,} \\.{5,}$", current_sex_df$year)

  current_sex_df = current_sex_df[keep_rows,]
  
  #add the field to each row
  current_sex_df$field = end_removed
  
  current_sex_df$year = substr(current_sex_df$year, start = 1, stop = 4) |> as.numeric()
 
  
  if (exists("fields_by_sex")){
    fields_by_sex = rbind(current_sex_df, fields_by_sex)
  }
  else {
    fields_by_sex = current_sex_df
  }
}



```

Here, we will read in our table of yearly breakdown
```{r}
sex_overall <- read_xls('/Users/zacharyclement/degree_sex_trends/all_degrees_by_sex.xls', skip = 4)

sex_overall <- sex_overall[,-c(7, 9, 11)]

cols_overall <- c("year", "assoc_t", "assoc_m", "assoc_f", "assoc_pct_f", "bach_t",
                  "bach_m", "bach_f", "bach_pct_f", "mast_t","mast_m",
                    'mast_f', 'mast_pct_f', 'doct_t', 'doct_m', 'doct_f', 
                  'doct_pct_f')

colnames(sex_overall) <- cols_overall

keep_rows <- grep(pattern = "^\\d{4}-\\d{2,}$", sex_overall$year) #this gets rid of the projected years, but that's okay



sex_overall <- sex_overall[keep_rows,]

sex_overall$year = substr(sex_overall$year, start = 0, stop = 4) 



```



Now, let's make some graphs:

```{r}
library(ggplot2)

sex_overall <- lapply(sex_overall, as.numeric) |> data.frame()

temp_field <- fields_by_sex$field
fields_by_sex <- lapply(fields_by_sex, as.numeric) |> data.frame()
fields_by_sex$field = temp_field

ggplot() + 
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$bach_pct_f, 
                col = "Bachelor's")) +
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$doct_pct_f, 
                col = "Doctor's")) +
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$mast_pct_f, 
                col = "Master's")) +
  labs(x = "Year", y = "Percent Female", col = '') + 
  geom_hline(mapping = aes(col = "50/50", yintercept = 50))

```


```{r}
ggplot() + 
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$bach_t, 
                col = "Bachelor's")) +
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$doct_t, 
                col = "Doctor's")) +
  geom_line(aes(x = sex_overall$year, 
                y = sex_overall$mast_t, 
                col = "Master's")) +
  labs(x = "Year", y = "Degrees Awarded", col = '', title = "Number of total degrees awarded")
```



```{r}
unique(fields_by_sex$field)
```

Let's define a function to look at the trends in one field

```{r}

plot_by_field <- function(field, level){
  only_field <- fields_by_sex[fields_by_sex$field == field,]
  
  pct_female <- only_field[, paste0(level, "_f")] / only_field[, paste0(level, "_t")] 
  ggplot() + 
    geom_line(aes(x = only_field$year, y = pct_female )) +
    labs(title = paste0("Percent Female Degrees at ", level, " level in ", field),
         x = "Year", 
         y = "Percent Female")
}

plot_by_field("business", "doct")


```

Now, let's expand that to look at multiple fields
```{r}
plot_by_field <- function(fields, level){
  short_level = tolower(level) |> substr(start = 0, stop = 4)
  
  
  x_min = 3000 #initial values should never end up being the actual min
  x_max = 0
  y_min = 100
  y_max = 0
  
  p <- ggplot()
  
  for (field in fields){
    only_field <- fields_by_sex[fields_by_sex$field == field,]
    
    only_field <- only_field[order(only_field$year),] #should already be ordered but just in case
  
  pct_female <- only_field[, paste0(short_level, "_f")] / only_field[, paste0(short_level, "_t")]
  #check if we need to update the mins/maaxes
  y_min = min(y_min, pct_female)
  y_max = max(y_max, pct_female)
  x_min = min(x_min, only_field$year)
  x_max = max(x_max, only_field$year)
  
  #using the aes_ forces eager evaluation
    p <- p + geom_line(aes_(x = only_field$year, y = pct_female, col = field))
  }
   

   p + labs(title = paste0("Percent Female Degrees at ", level, " level"),
            col = "Field",
         x = "Year", 
         y = "Percent Female") + xlim(x_min, x_max) + ylim(y_min, y_max)
}

plot_by_field(c( "psychology", "business", 'agriculture and natural resources', 'architecture and related services', 'computer and information sciences'), "Doctor's")
```



